---
title: "Stat 133: Data Aggregation"
author: "Gaston Sanchez"
output: html_document
---

-----

## Hands-on with Data Aggregation

Read the data `tents1.csv` from the github repository:
```{r}
library(readr)

df <- read_csv("https://raw.githubusercontent.com/gastonstat/stat133/master/datasets/tents1.csv")
```

Quick inspection
```{r}
str(df, vec.len = 2)
```


-----


## Average Price by Brand

Obtain the average price by brand using a __for loop__. The output should be in a vector `avg_prices` like this one:

```{r, eval=FALSE}
     big-agnes          kelty         marmot            msr 
      376.3136       252.8071       354.6000       439.9500
          nemo     north-face            rei sierra-designs 
      374.9500       399.0000       286.2955       294.4750
```

_Your loop_
```{r}
# fill in

```



-----

## Average Price, Weight, and Height by Bestuse

1. Obtain the average price, average weight, and average height by `bestuse` using a for loop. The output should be in a list `pwh_list` like this one:

```{r, eval=FALSE}
$price
Backpacking     Carcamping Mountaineering 
   325.3669       376.2560       470.1583

$weight
Backpacking     Carcamping Mountaineering 
   2030.339       6572.556       3568.333 

$height
Backpacking     Carcamping Mountaineering 
   107.4746       173.5600       107.5000 
```

_Your loop_:
```{r}
# fill in

```

-----

2. Obtain the average price, average weight, and average height by `bestuse` using a for loop but now the output should be in a data.frame `pwh_df` like this one:

```{r, eval=FALSE}
                  price   weight   height
Backpacking    325.3669 2030.339 107.4746
Carcamping     376.2560 6572.556 173.5600
Mountaineering 470.1583 3568.333 107.5000 
```

_Your loop_
```{r}
# fill in

```

-----

3. Obtain the same averages (price, weight, and height) by `bestuse` using a for loop but now the output should be in a data.frame `df_pwh` like this one:

```{r, eval=FALSE}
       Backpacking Carcamping Mountaineering
price     325.3669    376.256       470.1583
weight   2030.3390   6572.556      3568.3333
height    107.4746    173.560       107.5000 
```

```{r}
# write your for loop

```

-----

## Aggregating data with no loops

1. __Average Price by Brand__: Obtain the average price by brand using `tapply()`
```{r}
# average price by brand with tapply()

```

2. __Average Price by Brand__: Obtain the same average price by brand using the `by()` function (see the documentation and how to use it)

```{r}
# average price by brand with by()

```

3. __Averages by Bestuse__: Obtain the average price, weight and height by `bestuse` using `aggregate()`
```{r}
# average price, weight, height by `bestuse` with aggregate()

```

4. __Averages by Bestuse and Seasons__: Obtain the average price, weight and height by `bestuse` and `seasons` using `aggregate()`
```{r}
# aggregate()

```

-----

### Extra manipulations

This extra section is just for demo purposes. R is extremely flexible for manipulating data: it allows you to perform one task in several ways, and then play with the format of the outputs.

You can use `split()` to partition a dataset according to a factor (the output will be a list). For instance, we can split `df` in terms of `bestuse`:
```{r}
# split data by bestuse (output as list)
pwh_list <- split(df[ ,c('price', 'weight', 'height')], df$bestuse)
```

Having a list, we can use `lapply()` and pass `colMeans()` to get the averages of price, height and weight:
```{r}
# lapply and colMeans
pwh_avgs <- lapply(pwh_list, colMeans)
pwh_avgs
```

We can transform `pwh_avgs` into a tabular layout (a matrix) by calling row binding `rbind()` inside `do.call()`, like this:
```{r}
do.call('rbind', pwh_avgs)
```

In this case, we can also use column binding `cbind()` inside `do.call()` and get a different arrangement:
```{r}
do.call('cbind', pwh_avgs)
```